\documentclass{article}

\input{../packages}
% \input{../packagesMemoir}
\input{../macros}

%opening
\title{Physics: Quick reference}
\author{vishvAs vAsuki}

\begin{document}
\maketitle

See \cite{resnickHalliday}.

\part{Movement, forces, work, energy}
\section{Kinematics}
Displacement s: a vector. Mass m, time t. Speed, velocity $v = \frac{ds}{dt}$, momentum = mv.

\section{Mechanics}
Acceleration $a = \frac{dv}{dt}$.

\subsection{Force}
F = ma; so the rate of change of momentum. A vector too.

Forces occur in pairs: Action, reaction; when one body exerts a force on another, an equal force is exerted by that other body on it.

Conservation of momentum.

Normal forces, friction.

\subsection{Modelling}
Simplification, a body as a particle; free body diagrams. Center of mass; center of gravity.

\subsection{Circular motion}
Perpendicular forces, angular velocity $\omega$, relationship with v, Angular acceleration, torque. The wheel.

\section{Forces}
\subsection{Gravity}
\subsubsection{Newton's model}
The law of gravity: $F = k \frac{m_{1}m_{2}}{r^{2}}$. Usually accurate enough in daily life.

\paragraph*{Drawbacks}
There is instantaneous action at a distance. But, according to special theory of relativity, no information can be transmitted faster than the speed of light; but if newton's law was an accurate model, you would have a gravitational telegraph which can transmit information instantaneously.

\subsubsection{Curvature in space-time model}
Mass curves space-time. Objects move through this curved space-time.

\paragraph*{Gravitational waves}
An object osscilating in a small pace causes waves in space-time.

Evidence of existance: Massive stars revolving around each other moving closer to each other, loosing energy; matches theoretical prediction of energy dissipated as gravitational waves.

\subsubsection{Application of Newton model}
\paragraph*{Gravity exerted by spherical objects}
An important special case. If sphere is symmetric in density, it turns out that using point mass located at the center of the sphere is an equally accurate model.

\paragraph*{Object falling through air, close to planetary surface}
Model: Acceleration due to gravity is constant: relative change in r is insignificant.

Air resistance, terminal velocity.

\section{Energy}
Work $w = \int_{a}^{b}fds$ , energy, power. The law of conservation of energy. Potential and keinetic energy. Heat. Escape velocity.

\section{Thermodynamics}
Molecular motion. Thermodynamic equilibrium is an equivalence relation.

Closed thermodynamic system: Change in the internal energy U = (Enthalpy) Heat supplied + work done.

Work/ energy: PV = nRT. Engines: external combustion and internal combustion. PV curve; cyclical processes, Isothermic and adiabatic processes. Area of a cycle: No perfect engine law. Carnot cycle. Refridgeration and compressors. The total entropy of any isolated thermodynamic system tends to increase over time, approaching a maximum value. Definition of Absolute Zero temperature. Boltzmann distribution of energy at temperature T. Heat: Conduction, convection, radiation. Stefan Boltzmann law. Radiation and absorption spectra.

\section{Electricity}
Charge: attraction and repulsion. The inverse square law. Lines of force, flux. Find charge inside a surface: Gauss' law. Electric potential energy; electric potential. Potential contours, electric field. Potential difference and force.

\subsection{Magnetism}
The Hall effect. Magnetic field B as force per unit moving charge: $\textbf{F} = q\textbf{v}\times\textbf{B}$; Relative direction of B. Natural magnets: electric dipoles. Electromagnets. Construction of electric motors and mechanical electricity generators. AC and DC.

\section{Quantum mechanics}
Matter waves. Quanta: Photoelectric effect.

\section{Relativity}
Conservation of energy and mass.

\part{Modelling important situations}
\section{Periodic Motion}
Longitudonal and transverse waves. Frequency, amplitude. Equation for displacement about a point; relationship with circular motion.

\section{Hydrodynamics}
Incompressible fluids. Bernoulli's law relates density, gravity, fluid velocity. Flight of airplanes and helicopters. Buoyancy. Viscosity.

\section{Electric circuits}
Electric current, resistivity, conductance. Resistance in a wire related to resistivity, thickness and length: The molecular view. V=IR. Heat generated by resistance. Kirchoff's law of voltage and current conservation in loops: linear equations solved using linear algebra \cite{strang}. Capacitors. Capacitance: charge stored, potential difference; plate surface area, distance, dipole moment. Electric transmission using earth and a wire.

\section{Electronics}
Semiconductors. Silicon, germanium. Positive and negative doping.

Diodes: The np junction. .5V to .7V barrier. Forward bias. Reverse bias: breakdown after 5V.

Transistors: pnp and npn.

Operational Amplifiers.

\bibliographystyle{plain}
\bibliography{physics}

\end{document}
