\documentclass{sig-alternate}
\input{packages}
\input{macros}
\input{MACPdef}

%
\def\sharedaffiliation{%
\end{tabular}
\begin{tabular}{c}}
%
\begin{document}
\conferenceinfo{RecSys2010,} {September 26--30, 2010, Barcelona, Spain.} 
\CopyrightYear{2010}
\crdata{978-1-60558-906-0/10/09}
\clubpenalty=10000
\widowpenalty = 10000

\title{Affiliation Recommendation using Auxiliary Networks}
\numberofauthors{4}
\author{
\alignauthor Vishvas Vasuki
\alignauthor Nagarajan Natarajan
\and
\alignauthor Zhengdong Lu
\alignauthor Inderjit Dhillon
\sharedaffiliation
\affaddr{Department of Computer Science and ICES, University of Texas at Austin.}
}


\date{}

\maketitle

\begin{abstract}
\input{abstract}
\end{abstract}

\category{H.2.8}{Database applications - Data mining}{}
\terms{Algorithms, Experimentation}

\section{Introduction}
\label{Introduction}
\input{introduction}

\section{Models}
\label{Models}
\input{models}


\section{Related work}
\label{Related work}
\input{relatedWork}

\section{Experimental Evaluation}
\label{Experimental Evaluation}
\input{experimentalEvaluation}

\section{Conclusion and Future Work}
\label{Conclusion and Future Work}
\input{conclusionAndFutureWork}

\section{Acknowledgements}
We thank Prateek Jain and Berkant Savas for helpful discussions. We also thank Alan Mislove \cite{Mislove} for providing access to Orkut and Youtube datasets. Zhengdong Lu is supported by an ICES postdoctoral fellowship at UT Austin. This research was supported by NSF grants CCF-0916309 and IIS-0713142.

\bibliographystyle{plain}
\bibliography{references}

\begin{comment}
\begin{thebibliography}{10}

\bibitem{GroupFormation}
Lars Backstrom, Dan Huttenlocher, Jon Kleinberg, and Xiangyang Lan.
\newblock Group formation in large social networks: membership, growth, and
  evolution.
\newblock In {\em KDD '06: Proceedings of the 12th ACM SIGKDD International
  Conference on Knowledge Discovery and Data Mining}, pages 44--54, New York,
  NY, USA, 2006. ACM.

\bibitem{baderChew08}
Brett~W. Bader and Peter~A. Chew.
\newblock Enhancing multilingual latent semantic analysis with term alignment
  information.
\newblock In {\em COLING '08: Proceedings of the 22nd International Conference
  on Computational Linguistics}, pages 49--56, Morristown, NJ, USA, 2008.
  Association for Computational Linguistics.

\bibitem{ChenInfMax}
Wei Chen, Yajun Wang, and Siyu Yang.
\newblock Efficient influence maximization in social networks.
\newblock In {\em KDD '09: Proceedings of the 15th ACM SIGKDD International
  Conference on Knowledge Discovery and Data Mining}, pages 199--208, New York,
  NY, USA, 2009. ACM.

\bibitem{GoogleCFLatent}
Wen-Yen Chen, Jon-Chyuan Chu, Junyi Luan, Hongjie Bai, Yi~Wang, and Edward~Y.
  Chang.
\newblock Collaborative filtering for orkut communities: discovery of user
  latent behavior.
\newblock In {\em WWW '09: Proceedings of the 18th International Conference on
  World wide web}, pages 681--690, New York, NY, USA, 2009. ACM.

\bibitem{GoogleCCF}
Wen-Yen Chen, Dong Zhang, and Edward~Y. Chang.
\newblock Combinational collaborative filtering for personalized community
  recommendation.
\newblock In {\em KDD '08: Proceeding of the 14th ACM SIGKDD International
  Conference on Knowledge Discovery and Data Mining}, pages 115--123, New York,
  NY, USA, 2008. ACM.

\bibitem{HofmannPLSI}
Thomas Hofmann.
\newblock Probabilistic latent semantic indexing.
\newblock In {\em SIGIR '99: Proceedings of the 22nd annual International ACM
  SIGIR Conference on Research and development in Information Retrieval}, pages
  50--57, New York, NY, USA, 1999. ACM.

\bibitem{KleinbergInfluence}
David Kempe, Jon Kleinberg, and \'{E}va Tardos.
\newblock Maximizing the spread of influence through a social network.
\newblock In {\em KDD '03: Proceedings of the Ninth ACM SIGKDD International
  Conference on Knowledge Discovery and Data Mining}, pages 137--146, New York,
  NY, USA, 2003. ACM.
\vfill\eject

\bibitem{KleinbergLinkPred}
David Liben-Nowell and Jon Kleinberg.
\newblock The link prediction problem for social networks.
\newblock In {\em CIKM '03: Proceedings of the Twelfth International Conference
  on Information and Knowledge Management}, pages 556--559, New York, NY, USA,
  2003. ACM.

\bibitem{Mislove}
Alan Mislove, Massimiliano Marcon, Krishna~P. Gummadi, Peter Druschel, and
  Bobby Bhattacharjee.
\newblock Measurement and analysis of online social networks.
\newblock In {\em IMC '07: Proceedings of the 7th ACM SIGCOMM Conference on
  Internet Measurement}, pages 29--42, New York, NY, USA, 2007. ACM.

\bibitem{savas10c}
B.~Savas and I.~S. Dhillon.
\newblock Fast and accurate low rank approximation of massive graphs.
\newblock Technical report TR--10--18, Department of Computer Science, The
  University of Texas at Austin, 2010.

\bibitem{SinghCMF}
Ajit~P. Singh and Geoffrey~J. Gordon.
\newblock Relational learning via collective matrix factorization.
\newblock In {\em KDD '08: Proceeding of the 14th ACM SIGKDD International
  Conference on Knowledge Discovery and Data Mining}, pages 650--658, New York,
  NY, USA, 2008. ACM.

\bibitem{WeiLMF}
Wei Tang, Zhengdong Lu, and Inderjit~S. Dhillon.
\newblock Clustering with multiple graphs.
\newblock In {\em ICDM '09: Proceedings of the 2009 Ninth IEEE International
  Conference on Data Mining}, pages 1016--1021, Washington, DC, USA, 2009. IEEE
  Computer Society.

\bibitem{Coevolution}
Elena Zheleva, Hossam Sharara, and Lise Getoor.
\newblock Co-evolution of social and affiliation networks.
\newblock In {\em KDD '09: Proceedings of the 15th ACM SIGKDD International
  Conference on Knowledge Discovery and Data Mining}, pages 1007--1016, New
  York, NY, USA, 2009. ACM.

\end{thebibliography}
\end{comment}


\end{document}
