\documentclass{article}

\input{../packages}
\input{../packagesMemoir}
\input{../macros}

%opening
\title{Chemistry: Quick reference}
\author{vishvAs vAsuki}

\begin{document}
\maketitle

\part{Elements and the atomic theory}
\section{Atomic theory of matter}
All elements are combinations of around 130 elements in different proportions. Each element has a certain chemical property.

\section{Orbitals}
s1, s2, p1, p2, p3, q, r. s is spherical, p is octal-shaped. Hybrid orbitals sp1 etc..

\part{Physical structure of matter}
\section{Molecular structure}
\subsection{Notation}
(Lewis) Letters represent various atoms; Dots represent valence electrons, possibly shared. .O:H.

Or, can use 1/2/3 dashes to represent single/ double/ triple bonds.

\subsubsection{3-D structures}
Can use 'Newman projection' diagrams.


\part{Dynamics of chemical reactions}
\section{Low level view}
Chemical reacion involves making and breaking of bonds; which requires atoms to collide. The former usually has to happen first: this requires an activation energy. This can be supplied with heat, light etc..

\section{Grosser view}
Reactions can be endothermic or exothermic.

\part{Organic chemistry}
Hydrocarbons. Aliphatic vs aromatic compounds. Aliphats can be alkanes, alkenes or alkynes. Functional groups in hydrocarbons: -NH2 or =O-OH.


\part{Proteins}
\section{Protein structure}
A chain of amino acids. Amino acids are 20. $H_3N-C_\ga(R)(CH_3)-COOH$. $COO-H_2N$ is the peptide bond: Just the right strength for biological activity: not too strong, not too weak.

\section{Protein folding}
\tbc

% \bibliographystyle{plain}
% \bibliography{chemistry}

\end{document}
