\documentclass[11pt]{article}
% \usepackage[left=3cm,top=1.5cm,right=3cm,bottom=3cm]{geometry}
\usepackage{caption}
\input{packages}
\input{packagesGrants}
\input{macros}
%opening
% \title{Social and biological network analysis}
% \author{}
% \date{}

\begin{document}
% \maketitle
% \tableofcontents

\newpage
\section{Introduction and problem description}
\input{introduction}


\section{Biological data from different organisms}
\label{sec:dataDescription}
\input{dataDescription}

\section{Related research}
\input{priorResearch}

\section{Proposed research}
\label{sec:proposedResearch}
\input{proposedResearch}


\vspace{-3ex}
\section{Preliminary research and results}
\label{sec:preliminaryResearch}
\input{preliminaryResearch}

\vspace{-2ex}
\section{Project timeline}
For proper coordination, one needs to establish a priority and sequencing among the proposed research topics. We outline key priorities on a yearly basis.

\vspace*{-3ex}
\paragraph*{Year 1}
\begin{itemize}
 \item Theoretical work on the development of algorithms for network analysis and matrix completion.
 \item Develop visualization toolbox to explore gene-phenotype data sets and gene-gene networks.
 \item Implementations in Matlab of predictors based on graph proximity and latent factor models.
 \item Generating gene-phenotype connection predictions using the implemented models.
\end{itemize}

\vspace*{-3ex}
\paragraph*{Year 2}
\begin{itemize}
 \item Exploration of the use of probabilistic modelling and one-class classification to solve the problem.
 \item Explore and develop graphical model-based approaches.
 \item Further development of network analysis algorithms and recommender systems.
 \item Application to other problems in social network analysis.
\end{itemize}

\vspace*{-3ex}
\paragraph*{Year 3}
\begin{itemize}
 \item Wet-lab experiments to check gene-phenotype predictions.
 \item Explore various ways of combining multiple predictors to produce better predictions.
 \item Release software tools for use by biologists.
 \item Explore the use of so-far unused sources of information for gene-phenotype link prediction.
\end{itemize}

\section{Research Team}
The research will be conducted at UT Austin in the Dhillon lab as well as the Marcotte lab. PI Dhillon is an expert in data mining and machine learning, while Co-PI Marcotte is an expert in bioinformatics and is pioneering the proposed new approach in disease modeling\cite{newsMarcotteNY}. PI Dhillon and Co-PI Marcotte have collaborated previously on research papers \cite{isd:emm:ur:2002a} and an NSF ITR project.

\pagebreak
\bibliographystyle{plain}
\bibliography{zhengdong,references}

\end{document}
